Flute and kink instabilities in a dynamically twisted flux tube with anisotropic plasma viscosity

### 2021-12-02 paper.tex
Remove all \rs commands

In general, the stability of a cylindrical twisted magnetic flux tube is analysed using perturbations of the form
\begin{equation}
  \label{eq:kink_perturbation}
\xi(r, \theta, z) = \xi(r) e^{i(m\theta + kz)},
\end{equation}
where $m$ and $k$ are the wavenumbers \rs{in the $\theta$ and $z$
  directions, respectively}{in the azimuthal and axial directions,
  $\theta$ and $z$, respectively and $r$ is the radial coordinate in
  cylindrical polars}. 
The helical kink instability occurs for perturbations where $m=1, k\ne0$ and is the only instability of this form which is a body instability, that is it moves the entire body of the flux tube. Perturbations where $m>1$ are termed flute or interchange instabilities.

When the magnetic field is sheared, as in a twisted magnetic flux
tube, an interchange instability (such as the flute
instability) is confined to a surface where the peaks and troughs
follow the shear of the field. That is, the instability is confined to
the surface where the perturbation wavevector $(0, m/r, k)$ is
perpendicular to the direction of the field, \rs{}{known as the
  ``resonance surface''}. In an axisymmetric twisted
flux tube the resonance surface is located at a radius $r$
given by
\begin{equation}
  \label{eq:resonant_surface}
\frac{m}{r} B_{\theta}(r) + kB_z(r) \approx 0.
\end{equation}

@@

\rs{In contrast to the precise form of the equilibrium state and
  perturbation studied in~\citet{quinnEffectAnisotropicViscosity2020},
  this is an experiment where a system is driven and instabilities
  occur organically as a result of noise providing a random
  perturbation. As a result of the driving, the flux tube is not in
  static equilibrium. Consequently, the stability criterion and linear
  growth rate presented previously are useful only as a guide and for
  approximate comparison.}{The stability criterion
  \eqref{eq:suydams_criterion} and the linear
  growth rate approximation \eqref{eq:fluting_growth_rate2} are useful
  only as a guide and for 
  approximate analysis of the numerical simulations presented in this
  work.  The precise form of the equilibrium state and the
  perturbations needed for the validity of
  \eqref{eq:suydams_criterion} and \eqref{eq:fluting_growth_rate2}
  were used in~\citet{quinnEffectAnisotropicViscosity2020}.
  In contrast, in the experiments reported in the following the system
  is driven and instabilities occur spontaneously due to random
  perturbation. As a result of the driving, the flux tube is also not
  in static equilibrium initially. }

@@

We consider the
magnetohydrodynamic equations for the density $\rho$, plasma
velocity $\vec{u}$, pressure $p$, magnetic field $\vec{B}$ and
internal energy $\varepsilon$, \rs{}{in their non-dimensionalised
  visco-resistive form}
\begin{subequations}
  \label{eq:MHD}
  \begin{gather}
%\begin{equation}
\label{eq:mhda}
\frac{D\rho}{Dt} = - \rho \vec{\nabla} \cdot \vec{u},\\
%\end{equation}
%\begin{equation}
\rho\frac{D\vec{u}}{Dt} = -\vec{\nabla} p + \vec{\jmath} \times \vec{B} + \vec{\nabla} \cdot \ten{\sigma},\\
%\end{equation}
%\begin{equation}
\frac{D\vec{B}}{Dt} = (\vec{B} \cdot \vec{\nabla})\vec{u} - (\vec{\nabla} \cdot \vec{u})\vec{B} + \eta \nabla^2 \vec{B},\\
%\end{equation}
%\begin{equation}
\rho\frac{D\varepsilon}{Dt} = -p \vec{\nabla} \cdot \vec{u} + {Q}_{\nu} + {Q}_{\eta},%\\
%\label{eq:energy}
%\end{equation}
    \end{gather}
\end{subequations}
where $\eta$ is the non-dimensionalised resistivity,
$\vec{\jmath} = \nabla \times \vec{B}$ is the current
density and \rs{$\ten{\sigma}$ is the viscous stress tensor (described
  below). The}{the} terms ${Q}_{\nu} = \ten{\sigma} : \vec{\nabla}\vec{u}$
  and ${Q}_{\eta} = \eta | \vec{\jmath} |^2$ are viscous heating and
  Ohmic heating, respectively. The system is closed by the inclusion
  of the equation of state for an ideal gas 
\begin{equation}
\varepsilon = \frac{p}{\rho(\gamma - 1)},
\end{equation}
with the specific heat ratio is given by $\gamma = 5/3$.

\rs{The isotropic (or Newtonian) viscosity stress tensor is given by
\begin{equation}
  \label{eq:isotropic_viscous_tensor}
  \ten{\sigma}_{\text{iso}} = \nu \ten{W},
\end{equation}
where $\nu$ is the viscous transport parameter, generally referred to as the viscosity,
\begin{equation}
  \label{eq:rate_of_strain}
  \ten{W} = \nabla\vec{u} + (\nabla\vec{u})^T - \tfrac{2}{3}(\nabla \cdot \vec{u})\ten{I},
\end{equation}
is the rate of strain tensor, and $\ten{I}$ is the  $3\times 3$
identity matrix. We use an identical model of anisotropic viscosity to
that used in
previous work~\citep{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021},
 the switching model, first presented
 in~\citet{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017}. While
 this model is intended to revert to an isotropic model in the
 vicinity of magnetic null points, we effectively only make use of the
 anisotropic component, since there are no null points present in the
 simulations reported here at any point. To this
 end, we set the switching parameter (which controls the degree of
 viscous anisotropy) to $a_0 = 150$ in Lare3d's control, a value
 sufficiently large so that the anisotropic viscous stress tensor is given by
\begin{equation}
  \label{eq:switching_model}
\ten{\sigma}_{\text{swi}} = \nu \left[\frac{3}{2}(\ten{W}\vec{b}\cdot\vec{b}) \left( \vec{b} \otimes \vec{b} - \frac{1}{3}\ten{I} \right)\right],
\end{equation}
where $\vec{b}$ is the unit vector in the direction of the magnetic field. This tensor is identical to the strong field approximation of Braginskii~\citep{braginskiiTransportProcessesPlasma1965}. Full use of the switching model can be found in~\citet{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017,quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}.
}{}

\rs{}{Two different models for the viscosity stress tensor $\ten{\sigma}$
will be compared and contrasted in this study. The first model is the
conventional} isotropic (or Newtonian) viscosity stress tensor \rs{}{used in
the vast majority of the existing literature, so that,} 
\begin{equation}
  \label{eq:isotropic_viscous_tensor}
\ten{\sigma = }\ten{\sigma}_{\text{iso}} = \nu \ten{W},
\end{equation}
where $\nu$ is the viscous transport parameter, generally referred to as the viscosity,
\begin{equation}
  \label{eq:rate_of_strain}
  \ten{W} = \nabla\vec{u} + (\nabla\vec{u})^T - \tfrac{2}{3}(\nabla \cdot \vec{u})\ten{I},
\end{equation}
is the rate of strain tensor, and $\ten{I}$ is the  $3\times 3$
identity tensor. The second model, which is the one of actual interest, is the anisotropic viscosity stress
tensor given by
\begin{equation}
  \label{eq:switching_model}
\rs{}{\ten{\sigma} = \ten{\sigma}_\text{aniso}} = \nu \left[\frac{3}{2}(\ten{W}\vec{b}\cdot\vec{b}) \left( \vec{b} \otimes \vec{b} - \frac{1}{3}\ten{I} \right)\right],
\end{equation}
where $\vec{b}$ is the unit vector in the direction of the magnetic
field.
\rs{}{
Expression \eqref{eq:switching_model} is identical to the strong field approximation of the
general anisotropic viscosity tensor derived in~\citet{braginskiiTransportProcessesPlasma1965}.
Expressions \eqref{eq:isotropic_viscous_tensor} and \eqref{eq:switching_model} arise as asymptotic limits of the
more general switching model used in our earlier works 
\citep{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017,quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}
which, includes both isotropic and anisotropic contributions and can
switch gradually between them depending on the strength of the
magnetic field at a given spacio-temporal location. For example, in
the vicinity of a null point where the magnetic field becomes weak the
isotropic viscosity contribution becomes dominant in the switching
model. Switching between the two limit cases is not relevant in the
present study where the variations in the magnetic field are not
significantly large.}

@@

\rs{}{The following initial and boundary conditions are used to form a
magnetic flux tube and excite instabilities by dynamic twisting.} 
The magnetic field is prescribed as initially straight and uniform,
\begin{equation}
\vec{B} = (0, 0, 1)^{\text{T}},
\end{equation}
in a cube of size $[-2,2]^3$, with further test simulations run using an elongated domain of size $4\times4\times20$. Initially, the velocity is set
everywhere to $\vec{u} = \vec{0}$, the density to $\rho = 1$,
and the
internal energy to $\varepsilon = 8.67\times 10^{-4}$. This corresponds
to a typical coronal temperature of $10^6$ K and a plasma beta of $1.11 \times 10^{-4}$. At the
boundaries, the magnetic field, velocity, density, and energy are fixed to their
initial values \rs{}{and their derivatives normal to the
boundaries are set to zero}  except
where twisting velocity ``driver'', described
below, is prescribed.
\rs{The spatial
derivatives of these variables are also set to zero at the
boundaries.}{} \rs{The resolution is $512$ grid points per dimension,
comparable to the highest resolution kink instability studies
of~\cite{hoodCoronalHeatingMagnetic2009} or medium resolution studies
of~\cite{barefordShockHeatingNumerical2015}.}{}

@@

The flux rope is formed by prescribing a slowly accelerating, rotating flow at the upper $z$-boundary as
\begin{equation}
  \label{eq:null_twisting_profile}
  \vec{u} = u_0 u_r(r) u_t(t) (-y, x, 0)^T,
\end{equation}
where $u_r(r)$ describes the radial profile of the twisting motion in terms of the radius $r^2 = x^2 + y^2$,
\begin{equation}
  \label{eq:radial_twisting_function}
  u_r(r) = u_{r0}(1 + \tanh(1 - r_d r^2)),
\end{equation}
where $r_d$ controls the radial extent of the driver, $u_{r0}$ is a normalising factor, and $u_t(t)$ describes the imposed acceleration of the twisting motion,
\begin{equation}
  \label{eq:ramping_up_function}
  u_t(t) = \tanh^2(t/t_r),
\end{equation}
where the parameter $t_r$ controls the time taken to reach the final driver velocity $u_0$. The functions $u_r(r)$ and $u_t(t)$ are plotted in figure~\ref{fig:kink_driver}. At the lower boundary, the flow is in the opposite direction. This form of driver allows the system to be accelerated slowly enough that the production of disruptive shocks and fast waves is minimal. It is unavoidable that some waves are produced during the boundary acceleration, however these usefully provide a source of noise which eventually forms a perturbation.
\rs{}{
The driver} velocity is set to $u_0 = 0.15$, the normalising factor is $u_{r0} = 2.08$, and setting $r_d = 5$ corresponds to a driver constrained to $r<1$ and with a peak velocity at $r\approx 0.38$. The ramping time is set to $t_r = 2$, resulting in an acceleration from $0$ to $u_0$ over approximately $5$ Alfv\'en times. These driver parameters correspond to a peak rotational period of $T_R = 15.92$, the length of time taken for one full turn to be injected by a single driver. Both drivers result in twist being added at a rate of $2\pi$ every $7.96$ Alfv\'en times. The twist profile across the entire flux tube develops in such a way that by $t\approx 20$, it is qualitatively similar to those studied in~\citet{quinnEffectAnisotropicViscosity2020,hoodCoronalHeatingMagnetic2009,barefordShockHeatingNumerical2015} however the length of the flux tubes differs significantly. This configuration produces a $z$-directed tube of increasingly twisted magnetic field that eventually becomes unstable to both the flute instability and the helical kink instability.

\rs{}{The problem formulated above is solved numerically using the
  staggered-grid, Lagrangian–Eulerian remap code for 3D MHD
  simulations Lare3D of~\cite{arberStaggeredGridLagrangian2001} where
  a  new module for anisotropic viscosity has been
  included as detailed in~\citet{quinnKelvinHelmholtzInstabilityCollapse2021}.} The resolution used in the
current work is  $512$ grid points per dimension, comparable to the
highest resolution kink instability studies
of~\cite{hoodCoronalHeatingMagnetic2009} or medium resolution studies
of~\cite{barefordShockHeatingNumerical2015}.

@@

\rs{Two pairs of simulations were performed,}{We focus the attention
on two selected pairs of  
simulations, one pair where the background resistivity is set to
$\eta=10^{-3}$ and another where $\eta=10^{-4}$. As
in~\citet{quinnEffectAnisotropicViscosity2020}, only background
resistivity is used. Each pair consists of one simulation
using isotropic viscosity~\eqref{eq:isotropic_viscous_tensor} and
another one using the anisotropic model~\eqref{eq:switching_model}. The
value of viscosity is set to $\nu = 10^{-4}$ in all cases.} 
%
The overall development of both the flute and the kink
instabilities is broadly similar for the two values of resistivity and
is described  initially. Similar simulations were performed with a longer flux tube of length $20$ instead of the tubes shown here with length $4$, and the results were found to be qualitatively similar. Focus is 
then placed on the detailed description of instabilities in the
$\eta=10^{-4}$ cases, with the aim of comparing the effects of the two
viscosity models. \rs{These cases illustrate the development of the instabilities while showcasing the effect of the
viscosity models.}{} Then further features of the $\eta=10^{-3}$
cases are summarised. \rs{without a full exploration of the (qualitatively
similar) results.}{}

\subsection{\rs{Overview of instability development}{Mechanism and
    general development of instability}}

\begin{figure*}
  \centering
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{current_profiles.pdf}
      \caption{Current}
      \label{fig:current_profiles}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{ohmic_heating_profiles.pdf}
      \caption{Ohmic heating}
      \label{fig:kink_straight_ohmic_heating_profile}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{pressure_profiles.pdf}
      \caption{Pressure}
      \label{fig:pressure_profiles}
    \end{subfigure}
  \mycaption{Gradients in the current density generate pressure
    gradients through Ohmic heating.}{\rs{}{Axial current density (a), Ohmic heating
      (b) and pressure (c) as functions of the radial distance from
      the tube axis.}{} All plots are from anisotropic
    cases when $t=20$ through the plane $z=0$. Note the sign of the
    axial current density $j_z$ has been flipped for comparison and
    the Ohmic heating is given by $Q_{\eta} = \eta j^2$. The pressure
    profile of an additional test-case where $\eta=0$ is also
    shown. Line types are indicated in the legend.}%
  \label{fig:pressure_and_heating}
\end{figure*}

\rs{In all cases, the initial reaction to the twisting at the upper
and lower boundaries is two}{Initially and in all cases computed, the
twisting at the upper and lower boundaries gives rise to a pair of}
torsional Alfv\'en waves which proceed to travel along the tube
from the upper and lower boundaries to their respective
opposite boundaries. The interaction between these waves produces an 
oscillating pattern in the kinetic energy with a period of
approximately $4$ Alfv\'en times, equal to the time taken for an
Alfv\'en wave to traverse the entire length of the domain
as visible early in figure~\ref{fig:kink_ke-4}.

@@

In all cases unstable to the flute instability, it
occurs  between $t=20$ and $t=30$. \rs{During this
  time, the}{The} continued driving at the boundaries eventually
injects enough twist that the tube also becomes unstable to the kink
instability. This initially develops linearly alongside \rs{}{or shortly
after} the flute instability and then erupts during the
its  nonlinear phase, dominating the dynamics and
disrupting the   flute instability. The \rs{}{onset and
the competition} of the two instabilities is strongly affected by
the value of $\eta$ and the viscosity model used.

@@

\rs{}{We now describe the evolution and competition of flute and
  kink instabilities in case of resistivity $\eta=10^{-4}$.}
Figure~\ref{fig:kink_pressure_slices-4} shows the pressure profile
\rs{the viscosity model is switching for times}{of the \rs{switching}{anisotropic} viscosity case 
\eqref{eq:switching_model} at
time moments} $t=26$, $28$ and $30$ and at $z=0$. At $t=26$
the tube becomes unstable to  flute
instability with azimuthal wavenumber $m=4$, when the
plasma begins to  bulge out diagonally from the
high-pressure core as seen in figure~\ref{fig:swi-4_pressure_13}. As  
the bulges move radially outwards into lower pressure regions they
expand and accelerate, resulting in the entire pressure structure
appearing taking the shape of a four-leaf clover (figure~\ref{fig:swi-4_pressure_14}). By
$t=30$ the kink instability has disrupted the flute
instability and is developing nonlinearly
as evident in figure~\ref{fig:swi-4_pressure_15}. As is typical of nonlinear kink
development, the tube continues to release its stored potential energy
in the form of kinetic energy and heat and the contained plasma becomes highly
mixed. In the  isotropic viscosity case
which will not be illustrated by a separate figure, the
flute instability is present but damped, and \rs{}{it is
  quickly outcompeted by} the kink instability which quickly dominates the dynamics.

\begin{figure}
  \centering
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{field_line_plots/cropped/v-4r-4-isotropic_0014_cropped.png}
      \caption{Isotropic}
      \label{fig:field_line_plots_iso}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{field_line_plots/cropped/v-4r-4-switching_0014_cropped.png}
      \caption{Anisotropic}
      \label{fig:field_line_plots_swi}
    \end{subfigure}
\mycaption{Simultaneous development of flute and kink
  instabilities in the isotropic and anisotropic cases
  illustrated by field lines and pressure contours.}{Field
  lines and contours of pressure (where $p=0.3$) are plotted at
  $t=28$. Also shown is the velocity driver \rs{as a slice}{$u_r(\sqrt{x^2+y^2})$ at $z=2$}. The flute instability grows in both cases, though faster in the anisotropic case. The initial stages of the kink instability can also be observed in the field lines of the isotropic case in subfigure~\ref{fig:field_line_plots_iso}.}
\label{fig:kink_field_line_plots}%
\end{figure}

Figure~\ref{fig:kink_field_line_plots} shows the effect the viscosity
models have on the initial stages of the flute and kink
instabilities in 3D. While the flute instability is
observed in both cases, it is damped in the isotropic case and grows
faster in the anisotropic case. In the latter case,
the extended development of the flute instability
appears to disrupt the inner core of field lines and, \rs{(as shall be
  seen)}{as will be discussed further below,} slows the growth of the
kink instability. In the isotropic case, the flute
instability has been damped to the extent that the kink instability
grows uninhibited and quickly disrupts the fluting.

>>>>

In general, the stability of a cylindrical twisted magnetic flux tube is analysed using perturbations of the form
\begin{equation}
  \label{eq:kink_perturbation}
\xi(r, \theta, z) = \xi(r) e^{i(m\theta + kz)},
\end{equation}
where $m$ and $k$ are the wavenumbers in the azimuthal and axial directions,
  $\theta$ and $z$, respectively and $r$ is the radial coordinate in
  cylindrical polars. 
The helical kink instability occurs for perturbations where $m=1, k\ne0$ and is the only instability of this form which is a body instability, that is it moves the entire body of the flux tube. Perturbations where $m>1$ are termed flute or interchange instabilities.

When the magnetic field is sheared, as in a twisted magnetic flux
tube, an interchange instability (such as the flute
instability) is confined to a surface where the peaks and troughs
follow the shear of the field. That is, the instability is confined to
the surface where the perturbation wavevector $(0, m/r, k)$ is
perpendicular to the direction of the field, known as the
  ``resonance surface''. In an axisymmetric twisted
flux tube the resonance surface is located at a radius $r$
given by
\begin{equation}
  \label{eq:resonant_surface}
\frac{m}{r} B_{\theta}(r) + kB_z(r) \approx 0.
\end{equation}

@@

The stability criterion \eqref{eq:suydams_criterion} and the linear growth rate
approximation \eqref{eq:fluting_growth_rate2} are useful only as a guide and
for approximate analysis of the numerical simulations presented in this work.
The precise form of the equilibrium state and the perturbations needed for the
validity of \eqref{eq:suydams_criterion} and \eqref{eq:fluting_growth_rate2}
were used in~\citet{quinnEffectAnisotropicViscosity2020}. In contrast, in the
experiments reported in the following the system is driven and instabilities
occur spontaneously due to random perturbation. As a result of the driving, the
flux tube is also not in static equilibrium initially.

@@

We consider the
magnetohydrodynamic equations for the density $\rho$, plasma
velocity $\vec{u}$, pressure $p$, magnetic field $\vec{B}$ and
internal energy $\varepsilon$, in their non-dimensionalised
  visco-resistive form

\begin{subequations}
  \label{eq:MHD}
  \begin{gather}
%\begin{equation}
\label{eq:mhda}
\frac{D\rho}{Dt} = - \rho \vec{\nabla} \cdot \vec{u},\\
%\end{equation}
%\begin{equation}
\rho\frac{D\vec{u}}{Dt} = -\vec{\nabla} p + \vec{\jmath} \times \vec{B} + \vec{\nabla} \cdot \ten{\sigma},\\
%\end{equation}
%\begin{equation}
\frac{D\vec{B}}{Dt} = (\vec{B} \cdot \vec{\nabla})\vec{u} - (\vec{\nabla} \cdot \vec{u})\vec{B} + \eta \nabla^2 \vec{B},\\
%\end{equation}
%\begin{equation}
\rho\frac{D\varepsilon}{Dt} = -p \vec{\nabla} \cdot \vec{u} + {Q}_{\nu} + {Q}_{\eta},%\\
%\label{eq:energy}
%\end{equation}
    \end{gather}
\end{subequations}

where $\eta$ is the non-dimensionalised resistivity,
$\vec{\jmath} = \nabla \times \vec{B}$ is the current
density and the terms ${Q}_{\nu} = \ten{\sigma} : \vec{\nabla}\vec{u}$
  and ${Q}_{\eta} = \eta | \vec{\jmath} |^2$ are viscous heating and
  Ohmic heating, respectively. The system is closed by the inclusion
  of the equation of state for an ideal gas

\begin{equation}
\varepsilon = \frac{p}{\rho(\gamma - 1)},
\end{equation}
with the specific heat ratio is given by $\gamma = 5/3$.

Two different models for the viscosity stress tensor $\ten{\sigma}$
will be compared and contrasted in this study. The first model is the
conventional isotropic (or Newtonian) viscosity stress tensor used in
the vast majority of the existing literature, so that, 
\begin{equation}
  \label{eq:isotropic_viscous_tensor}
\ten{\sigma = }\ten{\sigma}_{\text{iso}} = \nu \ten{W},
\end{equation}
where $\nu$ is the viscous transport parameter, generally referred to as the viscosity,
\begin{equation}
  \label{eq:rate_of_strain}
  \ten{W} = \nabla\vec{u} + (\nabla\vec{u})^T - \tfrac{2}{3}(\nabla \cdot \vec{u})\ten{I},
\end{equation}
is the rate of strain tensor, and $\ten{I}$ is the  $3\times 3$
identity tensor. The second model, which is the one of actual interest, is the anisotropic viscosity stress
tensor given by
\begin{equation}
  \label{eq:switching_model}
\ten{\sigma} = \ten{\sigma}_\text{aniso} = \nu \left[\frac{3}{2}(\ten{W}\vec{b}\cdot\vec{b}) \left( \vec{b} \otimes \vec{b} - \frac{1}{3}\ten{I} \right)\right],
\end{equation}
where $\vec{b}$ is the unit vector in the direction of the magnetic
field.

Expression \eqref{eq:switching_model} is identical to the strong field approximation of the
general anisotropic viscosity tensor derived in~\citet{braginskiiTransportProcessesPlasma1965}.
Expressions \eqref{eq:isotropic_viscous_tensor} and \eqref{eq:switching_model} arise as asymptotic limits of the
more general switching model used in our earlier works 
\citep{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017,quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}
which, includes both isotropic and anisotropic contributions and can
switch gradually between them depending on the strength of the
magnetic field at a given spacio-temporal location. For example, in
the vicinity of a null point where the magnetic field becomes weak the
isotropic viscosity contribution becomes dominant in the switching
model. Switching between the two limit cases is not relevant in the
present study where the variations in the magnetic field are not
significantly large.

@@

The following initial and boundary conditions are used to form a
magnetic flux tube and excite instabilities by dynamic twisting. 
The magnetic field is prescribed as initially straight and uniform,
\begin{equation}
\vec{B} = (0, 0, 1)^{\text{T}},
\end{equation}
in a cube of size $[-2,2]^3$, with further test simulations run using an elongated domain of size $4\times4\times20$. Initially, the velocity is set
everywhere to $\vec{u} = \vec{0}$, the density to $\rho = 1$,
and the
internal energy to $\varepsilon = 8.67\times 10^{-4}$. This corresponds
to a typical coronal temperature of $10^6$ K and a plasma beta of $1.11 \times 10^{-4}$. At the
boundaries, the magnetic field, velocity, density, and energy are fixed to their
initial values and their derivatives normal to the
boundaries are set to zero  except
where twisting velocity ``driver'', described
below, is prescribed.

@@

The flux rope is formed by prescribing a slowly accelerating, rotating flow at the upper $z$-boundary as
\begin{equation}
  \label{eq:null_twisting_profile}
  \vec{u} = u_0 u_r(r) u_t(t) (-y, x, 0)^T,
\end{equation}
where $u_r(r)$ describes the radial profile of the twisting motion in terms of the radius $r^2 = x^2 + y^2$,
\begin{equation}
  \label{eq:radial_twisting_function}
  u_r(r) = u_{r0}(1 + \tanh(1 - r_d r^2)),
\end{equation}
where $r_d$ controls the radial extent of the driver, $u_{r0}$ is a normalising factor, and $u_t(t)$ describes the imposed acceleration of the twisting motion,
\begin{equation}
  \label{eq:ramping_up_function}
  u_t(t) = \tanh^2(t/t_r),
\end{equation}
where the parameter $t_r$ controls the time taken to reach the final driver velocity $u_0$. The functions $u_r(r)$ and $u_t(t)$ are plotted in figure~\ref{fig:kink_driver}. At the lower boundary, the flow is in the opposite direction. This form of driver allows the system to be accelerated slowly enough that the production of disruptive shocks and fast waves is minimal. It is unavoidable that some waves are produced during the boundary acceleration, however these usefully provide a source of noise which eventually forms a perturbation.

The driver velocity is set to $u_0 = 0.15$, the normalising factor is $u_{r0} = 2.08$, and setting $r_d = 5$ corresponds to a driver constrained to $r<1$ and with a peak velocity at $r\approx 0.38$. The ramping time is set to $t_r = 2$, resulting in an acceleration from $0$ to $u_0$ over approximately $5$ Alfv\'en times. These driver parameters correspond to a peak rotational period of $T_R = 15.92$, the length of time taken for one full turn to be injected by a single driver. Both drivers result in twist being added at a rate of $2\pi$ every $7.96$ Alfv\'en times. The twist profile across the entire flux tube develops in such a way that by $t\approx 20$, it is qualitatively similar to those studied in~\citet{quinnEffectAnisotropicViscosity2020,hoodCoronalHeatingMagnetic2009,barefordShockHeatingNumerical2015} however the length of the flux tubes differs significantly. This configuration produces a $z$-directed tube of increasingly twisted magnetic field that eventually becomes unstable to both the flute instability and the helical kink instability.

The problem formulated above is solved numerically using the
  staggered-grid, Lagrangian–Eulerian remap code for 3D MHD
  simulations Lare3D of~\cite{arberStaggeredGridLagrangian2001} where
  a  new module for anisotropic viscosity has been
  included as detailed in~\citet{quinnKelvinHelmholtzInstabilityCollapse2021}. The resolution used in the
current work is  $512$ grid points per dimension, comparable to the
highest resolution kink instability studies
of~\cite{hoodCoronalHeatingMagnetic2009} or medium resolution studies
of~\cite{barefordShockHeatingNumerical2015}.

@@

We focus the attention
on two selected pairs of  
simulations, one pair where the background resistivity is set to
$\eta=10^{-3}$ and another where $\eta=10^{-4}$. As
in~\citet{quinnEffectAnisotropicViscosity2020}, only background
resistivity is used. Each pair consists of one simulation
using isotropic viscosity~\eqref{eq:isotropic_viscous_tensor} and
another one using the anisotropic model~\eqref{eq:switching_model}. The
value of viscosity is set to $\nu = 10^{-4}$ in all cases. 
%
The overall development of both the flute and the kink
instabilities is broadly similar for the two values of resistivity and
is described  initially. Similar simulations were performed with a longer flux tube of length $20$ instead of the tubes shown here with length $4$, and the results were found to be qualitatively similar. Focus is 
then placed on the detailed description of instabilities in the
$\eta=10^{-4}$ cases, with the aim of comparing the effects of the two
viscosity models. Then further features of the $\eta=10^{-3}$
cases are summarised.

\subsection{Mechanism and
    general development of instability}

\begin{figure*}
  \centering
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{current_profiles.pdf}
      \caption{Current}
      \label{fig:current_profiles}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{ohmic_heating_profiles.pdf}
      \caption{Ohmic heating}
      \label{fig:kink_straight_ohmic_heating_profile}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
      \includegraphics[width=\linewidth]{pressure_profiles.pdf}
      \caption{Pressure}
      \label{fig:pressure_profiles}
    \end{subfigure}
  \mycaption{
  Gradients in the current density generate pressure gradients through Ohmic
  heating.}{Axial current density (a), Ohmic heating (b) and pressure (c) as
  functions of the radial distance from the tube axis. All plots are from
  anisotropic cases when $t=20$ through the plane $z=0$. Note the sign of the
  axial current density $j_z$ has been flipped for comparison and the Ohmic
  heating is given by $Q_{\eta} = \eta j^2$. The pressure profile of an
  additional test-case where $\eta=0$ is also shown. Line types are indicated
  in the legend.}%
  \label{fig:pressure_and_heating}
\end{figure*}

Initially and in all cases computed, the
twisting at the upper and lower boundaries gives rise to a pair of
torsional Alfv\'en waves which proceed to travel along the tube
from the upper and lower boundaries to their respective
opposite boundaries. The interaction between these waves produces an 
oscillating pattern in the kinetic energy with a period of
approximately $4$ Alfv\'en times, equal to the time taken for an
Alfv\'en wave to traverse the entire length of the domain
as visible early in figure~\ref{fig:kink_ke-4}.

@@

In all cases unstable to the flute instability, it
occurs  between $t=20$ and $t=30$. The continued driving at the boundaries eventually
injects enough twist that the tube also becomes unstable to the kink
instability. This initially develops linearly alongside or shortly
after the flute instability and then erupts during the
its  nonlinear phase, dominating the dynamics and
disrupting the   flute instability. The onset and
the competition of the two instabilities is strongly affected by
the value of $\eta$ and the viscosity model used.

@@

We now describe the evolution and competition of flute and
  kink instabilities in case of resistivity $\eta=10^{-4}$.
Figure~\ref{fig:kink_pressure_slices-4} shows the pressure profile
of the anisotropic viscosity case 
\eqref{eq:switching_model} at
time moments $t=26$, $28$ and $30$ and at $z=0$. At $t=26$
the tube becomes unstable to  flute
instability with azimuthal wavenumber $m=4$, when the
plasma begins to  bulge out diagonally from the
high-pressure core as seen in figure~\ref{fig:swi-4_pressure_13}. As  
the bulges move radially outwards into lower pressure regions they
expand and accelerate, resulting in the entire pressure structure
appearing taking the shape of a four-leaf clover (figure~\ref{fig:swi-4_pressure_14}). By
$t=30$ the kink instability has disrupted the flute
instability and is developing nonlinearly
as evident in figure~\ref{fig:swi-4_pressure_15}. As is typical of nonlinear kink
development, the tube continues to release its stored potential energy
in the form of kinetic energy and heat and the contained plasma becomes highly
mixed. In the  isotropic viscosity case
which will not be illustrated by a separate figure, the
flute instability is present but damped, and it is
  quickly outcompeted by the kink instability which quickly dominates the dynamics.

\begin{figure}
  \centering
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{field_line_plots/cropped/v-4r-4-isotropic_0014_cropped.png}
      \caption{Isotropic}
      \label{fig:field_line_plots_iso}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{field_line_plots/cropped/v-4r-4-switching_0014_cropped.png}
      \caption{Anisotropic}
      \label{fig:field_line_plots_swi}
    \end{subfigure}
\mycaption{Simultaneous development of flute and kink
  instabilities in the isotropic and anisotropic cases
  illustrated by field lines and pressure contours.}{Field
  lines and contours of pressure (where $p=0.3$) are plotted at
  $t=28$. Also shown is the velocity driver $u_r(\sqrt{x^2+y^2})$ at $z=2$. The flute instability grows in both cases, though faster in the anisotropic case. The initial stages of the kink instability can also be observed in the field lines of the isotropic case in subfigure~\ref{fig:field_line_plots_iso}.}
\label{fig:kink_field_line_plots}%
\end{figure}

Figure~\ref{fig:kink_field_line_plots} shows the effect the viscosity models
have on the initial stages of the flute and kink instabilities in 3D. While the
flute instability is observed in both cases, it is damped in the isotropic case
and grows faster in the anisotropic case. In the latter case, the extended
development of the flute instability appears to disrupt the inner core of field
lines and, as will be discussed further below, slows the growth of the kink
instability. In the isotropic case, the flute instability has been damped to
the extent that the kink instability grows uninhibited and quickly disrupts the
fluting.

\begin{figure}
  \centering
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{kinetic_energy-4.pdf}
      \caption{Kinetic Energy}
      \label{fig:kink_ke-4}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.49\textwidth}
      \includegraphics[width=\linewidth]{kinetic_energy_log-4.pdf}
      \caption{Growth rate estimation}
      \label{fig:kink_ke_log-4}
    \end{subfigure}
  \mycaption{Kinetic energy as a function of time showing the development and
  measured growth rates $\gamma$ and $\lambda$ of the flute and kink
  instabilities, respectively.}{Resistivity is $\eta=10^{-4}$ and the second
  plot is a enlarged version of the first.}
\label{fig:kink_str8_ke-4}%
\end{figure}

Despite the flute instability appearing in the isotropic
case (figure~\ref{fig:field_line_plots_iso}), only in the
anisotropic case can the onset of both the
flute and kink instabilities be seen in the kinetic
energy profile shown in figure~\ref{fig:kink_ke_log-4}. Here, the nonlinear
growth rates of the two instabilities are found to be $\gamma = 0.69$
for the flute and $\lambda = 2.55$ for the kink. The
onset times are approximately $t=27$ for the flute
instability and $t=29.5$ for
the kink. In the isotropic case, the growth rate of the kink, $\lambda
= 2.97$, is larger than in the anisotropic case,
although the onset times appear similar, and the kinetic energy
profile shows no evidence of flute instability growth.
